---
title: "Measuring Child Poverty Across Geographies"
subtitle: "Economics of Children — Data Exercise"
format:
  html:
    toc: true
    code-fold: show
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

Child poverty is not a single number — it depends enormously on *where* you draw
the line. The national child poverty rate hides huge differences between states;
the state rate hides huge differences between counties; the county rate hides
huge differences between neighborhoods.

In this exercise you'll pull real data from the U.S. Census Bureau's American
Community Survey (ACS) and measure child poverty at four geographic levels:
**nation, state, county, and census tract**. Each part ends with a "Your turn"
prompt — answer it in a text cell right below the code.

We'll use the `tidycensus` package (Kyle Walker), which queries the Census
Bureau's API and returns tidy data frames (and, if you ask for `geometry = TRUE`,
map-ready shapefiles too).

## Setup

You need a free Census API key: <https://api.census.gov/data/key_signup.html>

Run `census_api_key("YOUR KEY HERE", install = TRUE)` **once** in the R console
(not in this document) — it saves the key to your `.Renviron` so you don't have
to paste it into any file that might get shared.

```{r}
#| label: setup
# install.packages(c("tidycensus", "tidyverse", "tigris", "scales"))
library(tidycensus)
library(tidyverse)
library(tigris)
library(scales)

options(tigris_use_cache = TRUE)
```

## Part 1: The national child poverty rate

ACS table `S1701` ("Poverty Status in the Past 12 Months") breaks poverty down
by age. Variable `S1701_C03_002` is the **percent of children under 18 below
the poverty line**; `S1701_C03_001` is the percent for the *total* population,
which lets you compare child poverty to overall poverty.

```{r}
#| label: national
us_child_poverty <- get_acs(
  geography = "us",
  table = "S1701",
  year = 2022,
  survey = "acs1"
) |>
  filter(variable == "S1701_C03_002")

us_child_poverty
```

> **Your turn:** What was the U.S. child poverty rate in 2022? Pull
> `S1701_C03_001` as well — is child poverty higher or lower than poverty for
> the population overall? By how much?

## Part 2: State-by-state comparison

```{r}
#| label: states
state_child_poverty <- get_acs(
  geography = "state",
  table = "S1701",
  year = 2022,
  survey = "acs1"
) |>
  filter(variable == "S1701_C03_002") |>
  arrange(desc(estimate))

state_child_poverty |>
  slice(1:10) |>
  ggplot(aes(x = reorder(NAME, estimate), y = estimate)) +
  geom_col(fill = "#1f548a") +
  geom_errorbar(aes(ymin = estimate - moe, ymax = estimate + moe), width = 0.3) +
  coord_flip() +
  labs(
    title = "10 States with the Highest Child Poverty Rates",
    subtitle = "2022 American Community Survey, 1-Year Estimates",
    x = NULL, y = "Child poverty rate (%)"
  ) +
  theme_minimal()
```

> **Your turn:** Now find the 10 states with the *lowest* child poverty rates
> (try `slice_min()`, or flip the `arrange()`). What patterns do you notice —
> region, rural vs. urban mix, anything you know about state policy?

## Part 3: County-level variation within a state

The `S1701` table above used 1-year estimates, which the Census Bureau only
publishes for areas with 65,000+ people — too big a population floor for most
counties. For small geographies we switch to the 5-year estimates
(`survey = "acs5"`), which pool five years of data to be reliable at small
scales, at the cost of being less current.

```{r}
#| label: oregon-counties
oregon_counties <- get_acs(
  geography = "county",
  state = "OR",
  table = "S1701",
  year = 2022,
  survey = "acs5",
  geometry = TRUE
) |>
  filter(variable == "S1701_C03_002")

oregon_counties |>
  ggplot(aes(fill = estimate)) +
  geom_sf(color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1, labels = label_percent(scale = 1)) +
  labs(
    title = "Child Poverty Rate by County, Oregon",
    subtitle = "2018-2022 American Community Survey, 5-Year Estimates",
    fill = "Child poverty\nrate"
  ) +
  theme_void()
```

> **Your turn:** Which Oregon county has the highest child poverty rate? The
> lowest? Look up one of the high-poverty counties — what do you know about
> its economy or geography that might help explain the number? (Try swapping
> `state = "OR"` for your home state.)

## Part 4: Zooming in — census tracts within Multnomah County

```{r}
#| label: multnomah-tracts
multnomah_tracts <- get_acs(
  geography = "tract",
  state = "OR",
  county = "Multnomah",
  table = "S1701",
  year = 2022,
  survey = "acs5",
  geometry = TRUE
) |>
  filter(variable == "S1701_C03_002")

multnomah_tracts |>
  ggplot(aes(fill = estimate)) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(option = "magma", direction = -1, labels = label_percent(scale = 1)) +
  labs(
    title = "Child Poverty Rate by Census Tract, Multnomah County",
    subtitle = "2018-2022 American Community Survey, 5-Year Estimates",
    fill = "Child poverty\nrate"
  ) +
  theme_void()
```

> **Your turn:** Compare this tract-level map to the county-level map in Part
> 3. How much variation was "hidden" inside Multnomah County's single average?
> What does that suggest about the limits of using large geographic units to
> understand where poor children actually live?

## Reflection

Answer in a few sentences each:

1. As you zoom from nation → state → county → tract, does child poverty look
   more or less unequal? Why might that be?
2. Every ACS estimate comes with a margin of error (the `moe` column). Why
   does the margin of error tend to get *larger* as the geography gets
   smaller? What does that mean for how confidently you can say one tract has
   "higher" child poverty than a neighboring one?
3. Pick one geographic level from this exercise. What policy would you
   propose to reduce child poverty at that level, and why does the
   geographic scale matter for that policy's design?

## Notes

- Data: U.S. Census Bureau, American Community Survey, Table S1701.
- Package: `tidycensus` (Kyle Walker) — <https://walker-data.com/tidycensus/>
- 1-year ACS estimates are more current but only exist for geographies with
  65,000+ people; 5-year estimates cover every geography but are averaged
  over five years, so they lag and smooth out recent change.
