---
title: "Practice Problems 1.2"
format: html
---

Here are some practice problems to explore the penguins data set. First we need to load the penguin data set.

```{r}
library("palmerpenguins")

penguins
```

1.  What are the mean and standard deviation of bill length of the penguins?

```{r}
# create a variable that contains only the column for bill length



# mean 

# standard deviation

```

2.  What are the mean and standard deviation of the body mass of the penguins?

```{r}
# create a variable that contains only the column for body mass

# mean

# standard deviation


```

3.  What is the mean and median flipper length of the penguins?

```{r}
# create a variable that contains only the column for flipper length

# mean

# median

```

4.  How long are the largest flippers in this data set? How long is the shortest? (Hint: google how to find the minimum and maximum values in a vector). You can use the same variable that you created in 3!

```{r}
# longest flipper length

# shortest flipper length

```

5.  What is the range of bill depths of penguins? (Range is the maximum value - minimum value)

```{r}
# deepest bill depth

# shallowest bill depth

# range of bill depth
```

6.  How many species of penguins are in this data set (hint: there is a function to find distinct values in a vector)? List them using comments.

```{r}

```

7.  Find the value for the longest bill in the data set. Assign it to the variable longestBill.

```{r}

```

8.  Calculate the variance of body mass. (Hint: google variance in R)

```{r}

```

9.  Calculate the variance of flipper length.

```{r}

```

10. Calculate and compare the standard deviation of bill length and bill depth. Which has a larger standard deviation?

```{r}
# standard deviation of bill length

# standard deviation of bill depth

# Which has a larger standard deviation? 
```

11. Challenge: calculate the mean and standard deviation of bill length for gentoo penguins

```{r}




```
